A free/open-source Kazakh-Tatar machine translation system

### 2013-06-30 paper/paper.tex
some cramming; still not quite crammed enough

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a\label{ex:like-kaz}
\begingl
\gla Маған сөздік түзген ұнайды. //
\glb маған сөздік түз-GAн ұна-E-дI //
\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\a\label{ex:like-tat}
\begingl
\gla Мин сүзлек төзергә яратам. //
\glb мин сүзлек төз-IргA ярат-E-м //
\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:kerek}
\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:try}
\a
\begingl
\gla Айгүл оны табуға әрекет жасап жүр. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
\glft \eng{Aygül is trying to find him.} //
\endgl
\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
%\a
%\begingl
%\gla Мин син белән сөйлешергә килдем. //
%\glb мин син белән сөйле-ш-IргA кил-DI-м. //
%\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
%\glft \eng{I came to speak with you.} //
%\endgl
\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

\begin{figure*}[bhtp]
\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
%\begin{texttt}
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
}}%\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

@@

\begin{figure}
%	\subfloat[A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.]{
\begin{lstlisting}
<rule> 
  <match lemma="ақпарат" tags="n.attr"/> 
  <match lemma="құрал" tags="n.*.px3sp.*">
    <select lemma="чара"/>
  </match> 
</rule>
\end{lstlisting}
%}\\
%\subfloat[]{
%}
\todo{A couple more lexical-selection rules }
%\caption{Some lexical selection rules in Kazakh}
\caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
\label{fig:lrx}
%\vspace{-1em}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
  <rule>
    <match lemma="рухани" tags="adj"/>
    <match lemma="өмір" tags="n.*">
        <select lemma="тормыш"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects тормыш for өмір if preceded by the word рухани.}
\label{fig:lrx3}
\end{figure}

@@

{\bf Mor.\ disambiguation} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}\tag{\$} \^{}ол\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}\tag{\$} \\

>>>>

%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\a\label{ex:like-kaz}
%\begingl
%\gla Маған сөздік түзген ұнайды. //
%\glb маған сөздік түз-GAн ұна-E-дI //
%\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
%\glft \eng{I like writing dictionaries.} //
%\endgl
%\a\label{ex:like-tat}
%\begingl
%\gla Мин сүзлек төзергә яратам. //
%\glb мин сүзлек төз-IргA ярат-E-м //
%\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
%\glft \eng{I like writing dictionaries.} //
%\endgl
%\xe

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\a
\label{ex:like-kaz}
\begingl
\gla Маған сөздік түзген ұнайды. //
\glb маған сөздік түз-GAн ұна-E-дI //
\glc \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe
%\a\label{ex:like-tat}
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:like-tat}
\begingl
\gla Мин сүзлек төзергә яратам. //
\glb мин сүзлек төз-IргA ярат-E-м //
\glc \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe

@@

%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\label{ex:kerek}
%\a
%\begingl
%\gla Мен үйге қайтуым керек. //
%\glb мен үй-GA қайт-у-Iм керек //
%\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
%\glft \eng{I need to go home.} //
%\endgl
%\a
%\begingl
%\gla Миңа өйгә кайтырга кирәк. //
%\glb мин-GA өй-GA кайт-IргA кирәк //
%\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
%\glft \eng{I need to go home.} //
%\endgl
%\xe

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:kerek}
%\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\xe

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\label{ex:try}
%\a
%\begingl
%\gla Айгүл оны табуға әрекет жасап жүр. //
%\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
%\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
%\glft \eng{Aygül is trying to find him.} //
%\endgl
%\a
%\begingl
%\gla Айгөл аны табарга тырыша. //
%\glb Айгөл а-NI тап-AргA тырыш-E //
%\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
%\glft \eng{Aygöl is trying to find him.} //
%\endgl
%\xe

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:try}
%\a
\begingl
\gla Айгүл оны табуға әрекет жасап жүр. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
\glc Aygül he-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
\glft \eng{Aygül is trying to find him.} //
\endgl
\xe
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
%\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe
%\a
%\begingl
%\gla Мин син белән сөйлешергә килдем. //
%\glb мин син белән сөйле-ш-IргA кил-DI-м. //
%\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
%\glft \eng{I came to speak with you.} //
%\endgl
\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
%\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glc I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

\begin{figure*}[bhtp]
\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
%\begin{texttt}
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
}}%\end{texttt}
\vspace{-0.5em}
\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

@@

\begin{figure}
%	\subfloat[A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.]{
\begin{lstlisting}
<rule> 
  <match lemma="ақпарат" tags="n.attr"/> 
  <match lemma="құрал" tags="n.*.px3sp.*">
    <select lemma="чара"/>
  </match> 
</rule>
\end{lstlisting}
\vspace{-1em}
%}\\
%\subfloat[]{
%}
%\caption{Some lexical selection rules in Kazakh}
\caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
\label{fig:lrx}
%\vspace{-1em}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
  <rule>
    <match lemma="рухани" tags="adj"/>
    <match lemma="өмір" tags="n.*">
        <select lemma="тормыш"/>
    </match>
  </rule>
\end{lstlisting}
\vspace{-1em}
\caption{A lexical selection rule that selects тормыш for өмір if preceded by the word рухани.}
\label{fig:lrx3}
\end{figure}

@@

{\bf Mor.\ disambig.} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}\tag{\$} \^{}ол\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}\tag{\$} \\
